---
title: "Fire Detection"
format: html
---

### Introduction

Wildfires are serious natural hazard that can dramatically affect human and the ecosystems. Monitoring these events is essential for assessing ecological impact, supporting recovery planning, and improving future fire-management strategies. Satellite imagery provides an effective way to observe burned areas over large regions, even in remote or inaccessible locations.

In this notebook, we use Sentinel-2 Level-2A imagery to visualize and analyze a wildfire event by comparing images before and after the fire. By creating visual composites and calculating burn-related spectral indices, we can clearly identify the extent of damage and understand how the landscape has changed.

### What we will learn

-   🚀 Access and prepare Sentinel-2 L2A imagery for wildfire detection.
-   🛰️ Visualize pre-fire and post-fire scenes using true-color composite.
-   🛰️ Compare pre-fire and post-fire images to identify the affected areas..

### The case study

In this notebook, we focus on the same fire event examined in ["the EOPF Zarr in Action tutorial"](https://eopf-toolkit.github.io/eopf-101/61_sardinia_s2_tfci.html) a fire that occurred in the Province of Nuoro in the Italian region of Sardinia on June 10th, 2025.

### Import packages

```{r message=FALSE, warning=FALSE}
library(rstac)
library(terra)
library(tidyverse)
library(Rarr)
library(stars)
library(fs)
```


### Fixes to the Rarr package

```{r}
.url_parse_other <- function(url) {
  parsed_url <- httr::parse_url(url)
  bucket <- gsub(
    x = parsed_url$path, pattern = "^/?([[a-z0-9\\:\\.-]*)/.*",
    replacement = "\\1", ignore.case = TRUE
  )
  object <- gsub(
    x = parsed_url$path, pattern = "^/?([a-z0-9\\:\\.-]*)/(.*)",
    replacement = "\\2", ignore.case = TRUE
  )
  hostname <- paste0(parsed_url$scheme, "://", parsed_url$hostname)

  if (!is.null(parsed_url$port)) {
    hostname <- paste0(hostname, ":", parsed_url$port)
  }

  res <- list(
    bucket = bucket,
    object = object,
    region = "auto",
    hostname = hostname
  )
  return(res)
}

assignInNamespace(".url_parse_other", .url_parse_other, ns = "Rarr")

```

### Access Sentinel-2 L2A imagery for Pre and Post-fire event using Rstac

```{r}

# Time range for the pre-fire event (one week before the event)
pre_f  = "2025-06-03T01:00:00Z/2025-06-04T05:00:00Z"

# Time range for the post-fire event (about 10 days after the event)
post_f = "2025-06-21T01:00:00Z/2025-06-22T05:00:00Z"

# Bounding box (xmin, ymin, xmax, ymax) covering the burned area in Nuoro, Sardinia
search_bbox = c(8.847198, 40.193395, 8.938865, 40.24189)

```

```{r error=TRUE}
stac_source <- stac("https://stac.core.eopf.eodc.eu/")
stac_source |>
  get_request()

derive_store_array <- function(store, product_url) {
  store |>
    mutate(array = str_remove(path, product_url)) |>
    relocate(array, .before = path)
}

# Acccess the Pre-Fire image using rstac
pre_stac_query <- rstac::stac_search(
  q = stac_source,
  collections = "sentinel-2-l2a",
  bbox = search_bbox,
  datetime = pre_f,
  limit = 999
)
executed_pre_stac_query <- rstac::get_request(pre_stac_query)

# Read Pre-Fire product using product URL
pre_s2_l2a_product <- executed_pre_stac_query$features[[1]] |>
  assets_select(asset_names = "product")

pre_s2_l2a_product_url <- pre_s2_l2a_product |>
  assets_url()

pre_zarr_store <- pre_s2_l2a_product_url |>
  zarr_overview(as_data_frame = TRUE) |>
  derive_store_array(pre_s2_l2a_product_url)

# Quicklook for Pre-Fire RGB 
pre_zarr_store |>
  filter(array == "/quality/l2a_quicklook/r20m/tci") |>
  pull(path) |>
  read_zarr_array() |>
  aperm(c(2, 3, 1)) |>
  rast() |>
  plotRGB()
```

⛔ Unfortunately we were not successful when trying to access the post-fire image: invalid char in json text.

```{r error=TRUE}
# Acccess the Post-Fire image using rstac
post_stac_query <- rstac::stac_search(
  q = stac_source,
  collections = "sentinel-2-l2a",
  bbox = search_bbox,
  datetime = post_f,
  limit = 999
)
executed_post_stac_query <- rstac::get_request(post_stac_query)
executed_post_stac_query

# Read Post-Fire product using product URL
post_s2_l2a_product <- executed_post_stac_query$features[[1]] |>
  assets_select(asset_names = "product")

post_s2_l2a_product_url <- post_s2_l2a_product |>
  assets_url()

post_zarr_store <- post_s2_l2a_product_url |>
  zarr_overview(as_data_frame = TRUE) |>
  derive_store_array(post_s2_l2a_product_url)

# Quicklook for Post-Fire RGB 
post_zarr_store |>
  filter(array == "/quality/l2a_quicklook/r20m/tci") |>
  pull(path) |>
  read_zarr_array() |>
  aperm(c(2, 3, 1)) |>
  rast() |>
  plotRGB()
```